Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-09 paper/section-general-framework.tex
Split syn into exp and ty

The definition of relational environments is the most subtle part of
our framework. The straightforward approach is to assume an
interpretation of indexing expressions in which sorts are
%interpreted as 
sets and index operations are functions, and then define the
relational interpretation of a primtive type $\tyPrim{X}{e_1,...,e_n}$
in terms of the interpretations of $e_1,...,e_n$. This approach
suffices to prove free theorems and type isomorphisms. However, to
prove non-definability results, like the one in
\exref{ex:uninhabited-type}, we need a more refined interpretation
that accounts for the definable programs permitted by the algebraic
indexing. Here, we identify the properties we require of relational
environments for the abstraction theorem to hold. In
\autoref{sec:constr-rel-env}, we give two generic constructions of
relational environments.

@@

\begin{definition}
  \label{defn:valid-rel-env-family}
  An index context-indexed family of sets of relational environments
  $\{ \relEnv{E}(\Delta) \subseteq \mathrm{RelEnv}(\Delta) \}_\Delta$
  is \emph{substitutive} if:
  \begin{enumerate}
  \item it is closed under composition with simultaneous substitution:
    if $\rho \in \relEnv{E}(\Delta)$ and $\Delta \vdash \sigma
    \Rightarrow \Delta'$, then $\rho \circ \sigma \in
    \relEnv{E}(\Delta')$; and
%  \item If we have a pair of relation environments $\rho_1 \in
%    \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
%    \relEnv{E}(\Delta)$, along with a simultaneous substitution
%    $\Delta' \vdash \sigma \Rightarrow \Delta$, such that $\rho_1
%    \circ \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists
%    a $\rho \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho
%    \circ \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ
%    \pi_{i\mathord:s'} = \rho_2$.
  \item if $\rho_1 \in \relEnv{E}(\Delta', i\mathord:s)$ and $\rho_2
    \in \relEnv{E}(\Delta)$, and $\Delta \vdash \sigma \Rightarrow
    \Delta'$ is such that $\rho_1 \circ \pi_{i\mathord:s} = \rho_2
    \circ \sigma$, then there exists a $\rho \in \relEnv{E}(\Delta,
    i\mathord:s)$ such that $\rho \circ \sigma_{i\mathord:s} =
    \rho_1$ and $\rho \circ \pi_{i\mathord:s} = \rho_2$.
    % such that the outer edges of the
    % following diagram commute, for all primitive types $\tyPrimNm{X} \in
    % \mathit{PrimType}$:
    % % FIXME: check that these compositions are the right way round
    % % FIXME: re-do this as a set of equations
    % % FIXME: give a better English description of why this is needed
    % \begin{displaymath}
    %   \xymatrix{
    %   {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
    %   \\
    %   {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
    %   \\
    %   &
    %   &
    %   {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    % }
    % \end{displaymath}
    % Then there exists a relation environment $\rho \in
    % \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
    % two triangles in the bottom right of the diagram commute.
    % FIXME: mention that we already know that the square commutes
  \end{enumerate}
  A family $\relEnv{E}$ is \emph{pre-substitutive} if only the first
  point holds.
\end{definition}
%
% For the general
% framework, we assume that, for each index context $\Delta$, we are
% given a set of relation environments $\relEnv{E}(\Delta)$. Note that
% this assignment of sets of relation environments to index contexts is
% not assumed to be compositional: it is not necessarily the case that
% the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
% defined in terms of $\relEnv{E}(\Delta_1)$ and
% $\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
% properties:
%
\noindent
These conditions ensure that the relational interpretation of types we
define below behaves correctly with respect to simultaneous
substitution of index expressions (see
\lemref{lem:tyeqsubst-relational}, below).

@@

% The following lemma states that the relation interpretation of types
% that we have defined in this section behaves well: the first part of
% the lemma states that two types that are judgmentally equal are given
% equal relational interpretations, and the second part states that
% substitution of index expressions in types can be interpreted by the
% composition of relational environments with simultaneous
% substitutions.
\begin{lemma}\label{lem:tyeqsubst-relational}
\par
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
    \rsem{B}{\relEnv{E}}{\rho}$;
  \item If $\Delta' \vdash A \isType$ and either $\relEnv{E}$ is
    substitutive or $\relEnv{E}$ is pre-substitutive and $A$ is
    quantifier-free, then for all $\Delta \vdash \sigma \Rightarrow
    \Delta'$ and $\rho \in \relEnv{E}(\Delta)$,
    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
    \circ \sigma)$.
  \end{enumerate}
\end{lemma}
\noindent
Note that the equations in both parts of
\lemref{lem:tyeqsubst-relational} are well-typed by virtue of the
corresponding parts of \lemref{lem:tyeqsubst-erasure}.

@@

\begin{theorem}[Abstraction]\label{thm:abstraction}
  Assume $\relEnv{E}$ is substitutive. If $\Delta; \Gamma \vdash M :
  A$, then for all $\rho \in \relEnv{E}(\Delta)$ and $\eta_1, \eta_2
  \in \ctxtSem{\Gamma}$ such that $(\eta_1, \eta_2) \in
  \rsem{\Gamma}{\relEnv{E}}\rho$, we have $(\tmSem{M}\eta_1,
  \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.

  If $\relEnv{E}$ is pre-substitutive, then the above restricts to the
  case when $\Gamma$ only contains types of the form $\forall
  \vec{i}\mathord:\vec{s}. B$, where $B$ is quantifier-free, $A$ is
  quantifier-free and $M$ contains no uses of \TirName{UnivAbs},
  \TirName{ExPack} or \TirName{ExUnpack} and all uses of
  \TirName{UnivApp} are fully applied.
\end{theorem}

@@

We give two constructions of families $\relEnv{E}$ of sets of
relational environments that allow us to derive useful consequences of
\thmref{thm:abstraction}. To satisfy the hypothesis of
\thmref{thm:abstraction} we still need to prove, for each
instantiation of our general framework, that the program-level
primitive operations preserve all the relations in our
constructions. This is the content of Lemmas
\ref{lem:geom-environments-1}, \ref{lem:geom-environments-2},
\ref{lem:monoid-ops-related}, \ref{lem:environments-information-flow}
and \ref{lem:metric-environments}.

\paragraph{Model-based Relational Environments}
Our model-based relational environments are defined in terms of
models. To each sort $s \in \SortSet$ we assign a carrier set
$\semSort{s}$. We extend this assignment to index contexts and other
sequences of sorts (e.g., type arities) using the cartesian product of
sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of $e$. We further assume
that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
\semIndexExp{\cdot}_{\mathcal{M}})$ satisfying all the axioms in
$\IndexAxiomSet$.

@@

\paragraph{Definability Relational Environments}
We now construct relational environments that can be used for
non-definability results. To this end, consider the typing
$t\mathord:\SynTransl{2}; \Gamma \vdash M : \tyPrim{vec}{t + t} \to
\tyPrim{vec}{t}$, where $\Gamma$ contains the primitive operations on
vectors indexed by translations from \autoref{sec:affine-vector-ops}.
In \exref{ex:uninhabited-type} we claimed that this type is
uninhabited. Intuitively, it is clear why: with the operations on
vectors in \autoref{sec:affine-vector-ops}, there is no way of
cancelling the additional $t$ in the argument type to get the result
type. More generally, the vector operations only allow us to combine
indices according to the group operations, so if the index of the
result type of a function lies outside the free abelian group
generated by the indices of the input types, then we cannot write a
program of this type.  The key idea is to assign different relational
interpretations to types with indices not in some generated
collection of index expressions.

Given a set $S$ of index expressions that are all well-indexed in some
index context $\Delta$, define $\Gen(S)$ to be the set of terms that
are generated by the elements of $S$ and the primitive index
operations. We close under index expression equivalence to get the
sets $\Gen^\equiv(S) = \{ e \sepbar \exists s.\ \exists e' \in
\Gen(S).\ \Delta \vdash e \equiv e' : s \}$.

Relational environments are parameterised by sets of terms $S$:
\begin{displaymath}
  \rho^S\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
    \begin{array}{ll}
      \mathrm{Eq}_{\tyPrimSem{\tyPrimNm{X}}} & \textrm{if all}\ e_i \in \Gen^\equiv(S) \\
      R^\bullet_{\tyPrimNm{X}} & \textrm{otherwise}
    \end{array}
  \right.
\end{displaymath}
Thus we assign the equality relation to primitive types whose index
arguments are generated up to equivalence by $S$. For types whose
indices are not in $\Gen^\equiv(S)$, we require a binary relation
$R^\bullet_{\tyPrimNm{X}} \in \Rel(\tyPrimSem{\tyPrimNm{X}})$. We set
$\relEnv{E}_{R^\bullet}(\Delta) = \{ \rho^S \sepbar \forall e \in S.\
\exists s. \Delta \vdash e : s \}$.

We have not been able to prove that this family of sets of relational
interpretations is substitutive. Therefore our non-definability
results are subject to the quantifier-free conditions in
\thmref{thm:abstraction}.

% Hence we must make do with the
% following theorem:

% These relational environments are non-compositional: elements of
% $\relEnv{E}(\Delta,\Delta')$ are not purely constructed from elements
% of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a consequence,
% these relational environments will only be
% \emph{pre-substitutive}.

\begin{theorem}
  The families $\relEnv{E}_{R^\bullet}$ are pre-substitutive.
\end{theorem}

% FIXME: look forward to the applications

>>>>

The definition of relational environments is the most subtle part of
our framework.

%The straightforward approach is to assume an
%interpretation of indexing expressions in which sorts are
%interpreted as 
%sets and index operations are functions, and then define the
%relational interpretation of a primtive type $\tyPrim{X}{e_1,...,e_n}$
%in terms of the interpretations of $e_1,...,e_n$. This approach
%suffices to prove free theorems and type isomorphisms. However, to
%prove non-definability results, like the one in
%\exref{ex:uninhabited-type}, we need a more refined interpretation
%that accounts for the definable programs permitted by the algebraic
%indexing. Here, we identify the properties we require of relational
%environments for the abstraction theorem to hold. In
%\autoref{sec:constr-rel-env}, we give two generic constructions of
%relational environments.

@@

\begin{definition}
  \label{defn:valid-rel-env-family}
  An index context-indexed family of sets of relational environments
  $\{ \relEnv{E}(\Delta) \subseteq \mathrm{RelEnv}(\Delta) \}_\Delta$
  is \emph{substitutive} if:
  \begin{enumerate}
  \item it is closed under composition with simultaneous substitution:
    if $\rho \in \relEnv{E}(\Delta)$ and $\Delta \vdash \sigma
    \Rightarrow \Delta'$, then $\rho \circ \sigma \in
    \relEnv{E}(\Delta')$; and
%  \item If we have a pair of relation environments $\rho_1 \in
%    \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
%    \relEnv{E}(\Delta)$, along with a simultaneous substitution
%    $\Delta' \vdash \sigma \Rightarrow \Delta$, such that $\rho_1
%    \circ \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists
%    a $\rho \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho
%    \circ \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ
%    \pi_{i\mathord:s'} = \rho_2$.
  \item if $\rho_1 \in \relEnv{E}(\Delta', i\mathord:s)$ and $\rho_2
    \in \relEnv{E}(\Delta)$, and $\Delta \vdash \sigma \Rightarrow
    \Delta'$ is such that $\rho_1 \circ \pi_{i\mathord:s} = \rho_2
    \circ \sigma$, then there exists a $\rho \in \relEnv{E}(\Delta,
    i\mathord:s)$ such that $\rho \circ \sigma_{i\mathord:s} =
    \rho_1$ and $\rho \circ \pi_{i\mathord:s} = \rho_2$.
    % such that the outer edges of the
    % following diagram commute, for all primitive types $\tyPrimNm{X} \in
    % \mathit{PrimType}$:
    % % FIXME: check that these compositions are the right way round
    % % FIXME: re-do this as a set of equations
    % % FIXME: give a better English description of why this is needed
    % \begin{displaymath}
    %   \xymatrix{
    %   {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
    %   \\
    %   {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
    %   \\
    %   &
    %   &
    %   {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    % }
    % \end{displaymath}
    % Then there exists a relation environment $\rho \in
    % \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
    % two triangles in the bottom right of the diagram commute.
    % FIXME: mention that we already know that the square commutes
  \end{enumerate}
\end{definition}
%
% For the general
% framework, we assume that, for each index context $\Delta$, we are
% given a set of relation environments $\relEnv{E}(\Delta)$. Note that
% this assignment of sets of relation environments to index contexts is
% not assumed to be compositional: it is not necessarily the case that
% the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
% defined in terms of $\relEnv{E}(\Delta_1)$ and
% $\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
% properties:
%
\noindent
These conditions ensure that the relational interpretation of types we
define below behaves correctly with respect to simultaneous
substitution of index expressions (see
\lemref{lem:tyeqsubst-relational}, below).

@@

% The following lemma states that the relation interpretation of types
% that we have defined in this section behaves well: the first part of
% the lemma states that two types that are judgmentally equal are given
% equal relational interpretations, and the second part states that
% substitution of index expressions in types can be interpreted by the
% composition of relational environments with simultaneous
% substitutions.
\begin{lemma}\label{lem:tyeqsubst-relational}
\par
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
    \rsem{B}{\relEnv{E}}{\rho}$;
  \item If $\Delta' \vdash A \isType$ and $\relEnv{E}$ is
    substitutive, then for all $\Delta \vdash \sigma \Rightarrow
    \Delta'$ and $\rho \in \relEnv{E}(\Delta)$,
    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
    \circ \sigma)$.
  \end{enumerate}
\end{lemma}
\noindent
Note that the equations in both parts of
\lemref{lem:tyeqsubst-relational} are well-typed by virtue of the
corresponding parts of \lemref{lem:tyeqsubst-erasure}.

@@

\begin{theorem}[Abstraction]\label{thm:abstraction}
  Assume $\relEnv{E}$ is substitutive. If $\Delta; \Gamma \vdash M :
  A$, then for all $\rho \in \relEnv{E}(\Delta)$ and $\eta_1, \eta_2
  \in \ctxtSem{\Gamma}$ such that $(\eta_1, \eta_2) \in
  \rsem{\Gamma}{\relEnv{E}}\rho$, we have $(\tmSem{M}\eta_1,
  \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
\end{theorem}

@@

We give construction of families $\relEnv{E}$ of sets of
relational environments that allow us to derive useful consequences of
\thmref{thm:abstraction}. 
To satisfy the hypothesis of
\thmref{thm:abstraction} we still need to prove, for each
instantiation of our general framework, that the program-level
primitive operations preserve all the relations in our
constructions. This is the content of Lemmas
\ref{lem:geom-environments-1}, \ref{lem:geom-environments-2},
\ref{lem:monoid-ops-related}, \ref{lem:environments-information-flow}
and \ref{lem:metric-environments}.

%\paragraph{Model-based Relational Environments}
Our relational environments are defined in terms of
models. To each sort $s \in \SortSet$ we assign a carrier set
$\semSort{s}$. We extend this assignment to index contexts and other
sequences of sorts (e.g., type arities) using the cartesian product of
sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of $e$. We further assume
that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
\semIndexExp{\cdot}_{\mathcal{M}})$ satisfying all the axioms in
$\IndexAxiomSet$.

@@
